---
format: 
  closeread-html:
    remove-header-space: true
    cr-style:
      section-background-color: transparent
      narrative-background-color-sidebar: "#522180"
      narrative-text-color-sidebar: "#EFEFF8"
---

# Exploring the Iris Dataset

This story explores the famous Iris dataset using R 
and `ggplot2`.


:::{.cr-section layout="sidebar-left"}

:::{#cr-setup}
```{r}
#| warning: false
library(ggplot2)
library(gt)
library(quarto)
library(brand.yml)
plot_theme <- theme_brand_ggplot2("_brand.yml")
table_theme <- quarto::theme_brand_gt("_brand.yml")
color_palette <- list(
  discrete1 = c("#6E2DAB", "#49752E", "#FFB300", "#D7263D", "#008B8B")
)
theme_set(plot_theme + theme(
    plot.title = element_text(size = 15),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 13)
  ))
```

:::

Let's explore the Iris flowers dataset.
[@cr-iris]{scale-by="1.3"}

:::{#cr-iris}

```{r}
data(iris)

head(iris) |> 
  gt() |> 
  tab_header(
    title = md("**Iris Flowers**"),
    subtitle = "These are some beautiful species"
  ) |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  table_theme()
```
:::

The Iris dataset contains measurements of sepal 
length, sepal width, petal length, and petal width 
for 150 iris flowers, along with their species. It's
a classic dataset for classification and data 
exploration.

Let's start by visualizing the distribution of sepal
length for each species. We can see that *setosa* 
generally has shorter sepals than *versicolor* and 
*virginica*. [@cr-sepal-length]

:::{#cr-sepal-length}
```{r}
ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  geom_boxplot(alpha = 0.4) +
  labs(title = "Sepal Length by Species",
       y = "Sepal Length (cm)") + 
  scale_fill_manual(values = color_palette$discrete1) 
```
:::

Next, let's examine the petal length. Here, the 
differences between species are even more 
pronounced, with *setosa* having much shorter 
petals. [@cr-petal-length]

:::{#cr-petal-length}
```{r}
ggplot(iris, aes(x = Species, y = Petal.Length, fill = Species)) +
  geom_boxplot(alpha = 0.4) +
  labs(title = "Petal Length by Species",
       y = "Petal Length (cm)") + 
  scale_fill_manual(values = color_palette$discrete1)
```
:::

Now, let's look at the relationship between sepal 
length and sepal width, colored by species. We can 
observe some clustering, particularly for *setosa*. 
[@cr-sepal-scatter]

:::{#cr-sepal-scatter}
```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, fill = Species)) +
  geom_point(size = 5, alpha = 0.6, shape = 21, color = "white") +
  labs(title = "Sepal Length vs. Sepal Width by Species",
       x = "Sepal Length (cm)",
       y = "Sepal Width (cm)") +
  scale_fill_manual(values = color_palette$discrete1)
```
:::

Finally, let's visualize the relationship between 
petal length and petal width. This plot clearly 
shows three distinct clusters, making it easy to 
separate the species based on these two features. 
[@cr-petal-scatter]

:::{#cr-petal-scatter}
```{r}
ggplot(iris, aes(x = Petal.Length, y = Petal.Width, fill = Species)) +
  geom_point(size = 5, alpha = 0.6, shape = 21, color = "white") +
  labs(title = "Petal Length vs. Petal Width by Species",
       x = "Petal Length (cm)",
       y = "Petal Width (cm)") +
  scale_fill_manual(values = color_palette$discrete1)
```
:::

This exploration demonstrates how the Iris dataset, 
despite its simplicity, provides a great foundation 
for understanding data visualization and basic 
machine learning concepts.
:::
