---
title: "Predicting Iris Species with ML"
warning: false
message: false
format: html
---

## Load Packages
We begin by loading all the required packages for ML and visualization.


```{r}
#| label: setup
library(randomForest)
library(caret)
library(ggplot2)
library(dplyr)
library(tidyr)
library(gt)
# NEW
library(quarto)
library(brand.yml)
```

```{r}
#| label: setup2
#| code-fold: true

plot_theme <- theme_brand_ggplot2("_brand.yml")
table_theme <- quarto::theme_brand_gt("_brand.yml")
color_palette <- list(
  discrete1 = c("#6E2DAB", "#49752E", "#FFB300", "#D7263D", "#008B8B"), 
  discrete2 = c("#9B79CD", "#2F4F4F", "#B0C4DE", "#DAA520", "#A52A2A"), 
  sequential1 = c("#EFEFF8", "#C7BDE0", "#9B79CD", "#6E2DAB", "#522180"), 
  sequential2 = c("#EBF5E7", "#B8D3A8", "#82B471", "#49752E", "#30511F"), 
  sequential3 = c("#FFF8E1", "#FFDDA1", "#FFB300", "#CC8C00", "#996600"), 
  diverging1 = c("#A50F15", "#D7263D", "#F49C8F", "#FDF7EC", "#A5D6A7", "#49752E","#284A1A"), 
  diverging2 = c("#522180", "#6E2DAB", "#B6A0D9", "#FDF7EC", "#FFDDA1", "#FFB300","#CC8C00"))
theme_set(plot_theme + theme(
    plot.title = element_text(size = 15),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 13)
  ))
```

## Load Data

The next step is to load the `iris` dataset:

```{r}
#| label: data1
data(iris)
iris |> 
    head() |> 
    gt() |> 
    tab_header(title = "Iris Flowers") |> 
    table_theme()
```

## Create feature matrix and target vector

```{r}
#| label: data2
X <- iris[, 1:4]
y <- iris$Species
```

## Split the data (70% train, 30% test)

```{r}
#| label: data3
set.seed(42)
train_index <- createDataPartition(y, p = 0.7, list = FALSE)
X_train <- X[train_index, ]
X_test  <- X[-train_index, ]
y_train <- y[train_index]
y_test  <- y[-train_index]
```

## Standardize features

```{r}
#| label: standardization
# Calculate mean and sd manually for each column
scaling_params <- data.frame(
  feature = colnames(X_train),
  mean = apply(X_train, 2, mean),
  sd = apply(X_train, 2, sd)
)

# Display scaling parameters
scaling_params |> 
  gt() |> 
  tab_header(title = "Scaling Parameters") |>
  fmt_number(columns = c(mean, sd), decimals = 4) |> 
  table_theme()

# Apply scaling
X_train_scaled <- scale(X_train, center = scaling_params$mean, scale = scaling_params$sd)
X_test_scaled <- scale(X_test, center = scaling_params$mean, scale = scaling_params$sd)
```

## Train a Random Forest classifier

```{r}
#| label: train
set.seed(42)
model <- randomForest(x = X_train_scaled, 
                     y = y_train, 
                     ntree = 100, 
                     importance = TRUE)
```

## Make predictions

```{r}
#| label: predict
y_pred <- predict(model, X_test_scaled)
```

## Evaluate the model

```{r}
#| label: evaluate
accuracy <- sum(y_pred == y_test) / length(y_test)
cat("Model Accuracy:", sprintf("%.2f%%", accuracy * 100), "\n")
```

## Classification Report

```{r}
#| label: report
cat("Classification Report:\n")
conf_matrix <- confusionMatrix(y_pred, y_test)
conf_matrix 
```

## Confusion Matrix for Plotting

```{r}
#| label: confusion_matrix
cm <- table(Actual = y_test, Predicted = y_pred)
cm_df <- as.data.frame(cm)
```

## Plot 1: Confusion Matrix

```{r}
#| label: confusion_plot
confusion_plot <- ggplot(cm_df, aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1.5) +
  geom_text(aes(label = Freq), color = "white", size = 12, fontface = "bold") +
  labs(title = "Confusion Matrix", 
       x = "Predicted Species", 
       y = "Actual Species",
       fill = "Count") + 
  scale_fill_gradientn(colors = color_palette$sequential1) # NEW

confusion_plot
```

## Feature importance

```{r}
#| label: feature_importance
feature_importance <- data.frame(
  feature = rownames(importance(model)),
  importance = importance(model)[, "MeanDecreaseGini"]
) %>%
  arrange(desc(importance))

feature_importance |> 
  gt() |> 
  tab_header(title = "Feature Importance") |>
  fmt_number(columns = importance, decimals = 4) |> 
  table_theme()
```


## Plot 2: Feature Importance

```{r}
#| label: importance_plot
importance_plot <- ggplot(feature_importance, 
                         aes(x = reorder(feature, importance), y = importance)) +
  geom_col(alpha = 0.8, fill = color_palette$discrete1[1]) +
  coord_flip() +
  labs(title = "Feature Importance in Random Forest Model",
       x = "Feature",
       y = "Importance Score")  +
  theme(plot.title = element_text(hjust = 0.5))

importance_plot
```

## Create predictions DataFrame for visualization

```{r}
#| label: pred_df
test_results <- as.data.frame(X_test)
colnames(test_results) <- colnames(X)
test_results$actual <- y_test
test_results$predicted <- y_pred
test_results$correct <- test_results$actual == test_results$predicted
```

## Plot 3: Prediction Results (using first two features)

```{r}
#| label: prediction_plot
prediction_plot <- ggplot(test_results, 
                         aes(x = Sepal.Length, y = Sepal.Width, 
                             color = predicted, shape = actual)) +
  scale_color_manual(values = color_palette$discrete1) +
  geom_point(size = 4, alpha = 0.7) +
  labs(title = "Model Predictions on Test Set",
       x = "Sepal Length (cm)",
       y = "Sepal Width (cm)",
       color = "Predicted",
       shape = "Actual") +
  theme(plot.title = element_text(hjust = 0.5))

prediction_plot
```

## Save the model and preprocessing parameters to disk

```{r}
saveRDS(model, "iris_rf_model.rds")
saveRDS(scaling_params, "iris_scaler.rds")
```